---
title: "brain_gwas"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes for brain gwas overlaps

Overlap variants from with haqers:
Makowski et al 2022: Discovery of genomic loci of the human cerebral cortex using genetically informed brain atlases
 
Notes from the text
- They used snps in ld to assign snps to genes: All SNPs in LD (R2 > 0.6) with the 393 regional variants were mapped to genes using posi-


## Exploration of variants

```{r}
brain_gwas_snp_stats <- read.table("../data_raw/brain_gwas/makowski2022/S2-Table 1.csv",skip = 1,header = T,sep = ",")

```

First question:

Are any of these SNPs already in the GWAS catalog. If so, I have made the overlaps already.

```{r}
library(tidyverse)
gwas_catalog_snps <- read_tsv("../data_raw/gwas_catalog/gwas_catalog_v1.0.2-associations_e105_r2021-12-21.tsv")

length(unique(gwas_catalog_snps$MAPPED_TRAIT))

nrow(gwas_catalog_snps[,c("SNPS","MAPPED_TRAIT")])
nrow(unique(gwas_catalog_snps[,c("SNPS","MAPPED_TRAIT")]))

```

```{r}
n_brain_and_gwas_catalog <- sum(unique(brain_gwas_snp_stats$SNP) %in% unique(gwas_catalog_snps$SNPS))
```

Second question: 

Is any of the brain gwas snps in the gwas catalog snps that are already overlapped to haqers?
```{r}
haqers_olap_pheno_df <- readRDS("../data_output/rds/haqers_olap_pheno_df.Rds")

n_haqer_brain_overlaps <- sum(unique(haqers_olap_pheno_df$SNPS) %in% unique(brain_gwas_snp_stats$SNP))
```


```{r}
gwas_ld_snp_random_pheno_df_list <- readRDS("../data_output/rds/gwas_ld_snp_random_pheno_df_list.Rds")

```

```{r}
n_random_brain_overlaps <- unlist(lapply(gwas_ld_snp_random_pheno_df_list, function(x) sum(unique(x$SNPS) %in% unique(brain_gwas_snp_stats$SNP))))

```

```{r}
enframe(n_random_brain_overlaps) %>%
  ggplot(aes(value)) +
  geom_histogram() +
  geom_vline(xintercept = n_haqer_brain_overlaps, lty = "dashed", color = "red")
```
```{r}
( sum( n_random_brain_overlaps >= n_haqer_brain_overlaps) + 1 ) / 1001
```

