---
title: "gwas_catalog_exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

# Read gwas catalog data

```{r}
gwas_catalog_ancestry <- read_tsv("../data_raw/gwas_catalog/gwas_catalog-ancestry_r2021-12-21.tsv")
```

# Exploration of ancestries
```{r}
gwas_catalog_ancestry <- gwas_catalog_ancestry %>%
  mutate(broad_ancestral_category = `BROAD ANCESTRAL CATEGORY`) 

```

How many studies per ancestry are there for the top 10 most abundant single ancestry studies?
```{r}
gwas_catalog_ancestry %>%
  count(broad_ancestral_category,name = "ancestry_count") %>%
  top_n(n = 10) %>%
  mutate(broad_ancestral_category = fct_reorder(broad_ancestral_category,ancestry_count)) %>%
  ggplot(aes(broad_ancestral_category,ancestry_count, fill = broad_ancestral_category)) +
  geom_col() +
  coord_flip() +
  ylab(NULL) +
  theme_minimal() +
  theme(aspect.ratio = 1,legend.position = "none") +
  scale_fill_brewer(type = "qual",palette = 3)
```

How many studies per ancestry are there for the top 10 most abundant single ancestry studies, taking out European?

```{r}
gwas_catalog_ancestry %>%
  count(broad_ancestral_category,name = "ancestry_count") %>%
  top_n(n = 10) %>%
  mutate(broad_ancestral_category = fct_reorder(broad_ancestral_category,ancestry_count)) %>%
  filter(broad_ancestral_category != "European") %>%
  ggplot(aes(broad_ancestral_category,ancestry_count, fill = broad_ancestral_category)) +
  geom_col() +
  coord_flip() +
  ylab(NULL) +
  theme_minimal() +
  theme(aspect.ratio = 1,legend.position = "none") +
  scale_fill_brewer(type = "qual",palette = 3)
```

# Explore SNP data

```{r}
gwas_catalog_snp <- read_tsv("../data_raw/gwas_catalog/gwas_catalog_v1.0.2-associations_e105_r2021-12-21.tsv")
```

```{r}
gwas_catalog_snp %>%
  count(MAPPED_TRAIT,name = "trait_count") %>%
  top_n(n = 10) %>%
  mutate(MAPPED_TRAIT = fct_reorder(MAPPED_TRAIT,trait_count)) %>%
  ggplot(aes(MAPPED_TRAIT,trait_count, fill = MAPPED_TRAIT)) +
  geom_col() +
  coord_flip() +
  ylab(NULL) +
  theme_minimal() +
  theme(aspect.ratio = 1,legend.position = "none") +
  scale_fill_brewer(type = "qual",palette = 3)
```



There are multiple ancestries per study!
```{r}
rowSums(table(gwas_catalog_ancestry$`STUDY ACCESSION`,gwas_catalog_ancestry$broad_ancestral_category)) %>%
  enframe(name = "study_accession",value = "count") %>%
  ggplot(aes(count)) +
  geom_histogram()
  
```

For now focus on single ancestry studies
```{r}
!duplicated(gwas_catalog_ancestry$`STUDY ACCESSION`,)

gwas_catalog_ancestry_nodups <- gwas_catalog_ancestry[!(duplicated(gwas_catalog_ancestry$`STUDY ACCESSION`) | duplicated(gwas_catalog_ancestry$`STUDY ACCESSION`, fromLast = TRUE)),]

```

Join SNPs and Ancestries
```{r}
gwas_catalog_snp <- gwas_catalog_snp %>%
  select("CHR_ID","CHR_POS","SNP_ID_CURRENT","DISEASE/TRAIT","STUDY ACCESSION")

gwas_catalog_ancestry_nodups <- gwas_catalog_ancestry_nodups %>%
  select("broad_ancestral_category","ADDITONAL ANCESTRY DESCRIPTION","STUDY ACCESSION")

gwas_snp_ancestry<- left_join(gwas_catalog_snp, gwas_catalog_ancestry_nodups,by = "STUDY ACCESSION")

```

# Quick exploration with HAQERs

Read haqers
```{r}
library(rtracklayer)
haqers <- import.bed("../data_raw/bed/HAQER.bed")
```



Get SNPs as GRanges
```{r}
gwas_snp_ancestry$CHR_POS <- as.numeric(gwas_snp_ancestry$CHR_POS)

gwas_snp_ancestry <- gwas_snp_ancestry[!is.na(gwas_snp_ancestry$CHR_POS),]

gwas_snp_ancestry_granges <- GRanges(seqnames = gwas_snp_ancestry$CHR_ID,
        ranges = IRanges(start = gwas_snp_ancestry$CHR_POS,width = 1))

elementMetadata(gwas_snp_ancestry_granges) <- DataFrame(gwas_snp_ancestry)
```


```{r}
library(plyranges)
```

```{r}
seqlevels(haqers) <- str_remove(seqlevels(haqers),"chr")


haqers$name <- paste0("haqer",1:length(haqers))
haqers_snp_ancestry <- plyranges::join_overlap_inner_within(gwas_snp_ancestry_granges, haqers)
```
```{r}
haqers_snp_ancestry %>%
  as.data.frame() %>%
  ggplot(aes(DISEASE.TRAIT)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
```{r}
haqers_snp_ancestry %>%
  as.data.frame() %>%
  ggplot(aes(broad_ancestral_category)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()

```

```{r}

A <- GRanges(seqnames = "chr1",
        ranges = IRanges(10,width = 7))
B <- GRanges(seqnames = "chr1",
        ranges= IRanges(5,15))
join_overlap_inner_within(A, B)

```


```{r}
countOverlaps(haqers,gwas_snp_ancestry_granges) %>%
  enframe() %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 5)
```

```{r}
haqers_dist_to_nearest_snp <- distanceToNearest(haqers,gwas_snp_ancestry_granges)

haqers_dist_to_nearest_snp@elementMetadata$distance %>%
  enframe() %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 100)


summary(haqers_dist_to_nearest_snp@elementMetadata$distance < 100)
```


